The Key to a Data Parallel Compiler

### 2016-05-08 2016-spring-array/2016-array-key.tex
#1: Incorporate some fixes

We do not have a large scale understanding of the performance characteristics 
of these techniques in real life outside of daily use of the compiler.
Further analysis is
required to understand what sort of compiler optimizations and analyses are 
necessary to compile the Co-dfns compiler efficiently for the GPU and CPU. 
We currently lack a good comparison against existing techniques for the same 
language, since the current compiler is relatively unique in that no other 
compilers to our knowledge compile the same language to the same targets 
with the same sorts of optimizations. We intend to implement other paradigms 
and other languages which do have existing compiler implementations to compare 
relative performance on multiple platforms.

@@

The J programming language \cite{hui2014key} was the first practical,
general-purpose programming language to introduce the Key operator as a
primitive operator with the presumption of its general usefulness. The
rank operator (\verb;⍤;) used throughout this treatment also derives from
the J traditional, receiving particular interest throughout the APL
community \cite{bernecky1987rank,hui1995rank}.

@@

The APEX compiler \cite{bernecky1997apex} developed vectorized approaches to handling
certain analyses to compile traditional APL, including a SIMD tokenizer
\cite{bernecky2003tokenizer}.
 It uses a SSA representation, and converts the dynamic scope of
traditional APL functions into a static form early on. It also uses
a matrix format to represent the AST. Traditional APL did not have
nested function definitions, however, and thus the APEX compiler does
not have any specific approaches to dealing with function lifting.

Bernecky further identified methods of reducing or optimizing the
computational complexity or cost of certain array operations, allowing
improved performance of easy to understand array expressions
\cite{bernecky1999reducing}.

Timothy Budd implemented a compiler \cite{budd1984apl,budd2012apl} for APL which targeted
vector processors as well as one for C code. They used a method of
lazy evaluation to avoid intermediate data copying. Budd provided
thoughts and some ideas on how the compiler might be implemented in
parallel as well.

Walter Schwarz implemented an APL to C compiler for the ACORN system
targeting the CM-2 machine, demonstrating performance potential for
APL as a massively parallel language \cite{schwarz1991acorn}.

W. Ching and D. Ju have spent significant work on the ELI language
and other APL-class language
 implementations, especially on parallelized code and optimization. 
\cite{ching2000design,ching1994experimental,ching1993primitive,
ching1990automatic,hendriks1990sparse,ju1991exploitation,
ju1991performance}

>>>>

We do not have a large scale understanding of the performance characteristics 
of these techniques in real life outside of daily use of the compiler.
Further analysis is
required to understand what sort of compiler optimizations and analyses are 
necessary to compile the Co-dfns compiler efficiently for the GPU and CPU. 
We currently lack a good comparison against existing techniques both for 
the compiler design and the generated code. Some existing compilers are able 
to compile variants of APL to target the GPU and the CPU, but there are subtle 
differences that require care when comparing generated code, and the 
architecture of these compilers is often quite different. Comparing the 
relative performance both of the compiler designs (human factors and execution 
speed) and generated code is important to understanding these techniques more 
fully.

@@

Iverson introduced the idea of the Rank operator while at Sharp Associates 
\cite{iverson1983rationalized} as a part of ``Rationalized APL.'' 
The rank operator (\verb;⍤;) used here derived from the J language, a 
continuation of the rationalized APL idea, with many ideas such as rank 
percolating back into APL implementations. \cite{bernecky1987rank,hui1995rank}
The J programming language \cite{hui2014key} was the first practical,
general-purpose programming language to introduce the Key operator as a
primitive operator with the presumption of its general usefulness.

Bernecky \cite{bernecky2015abstract} argues that the increased programmability 
and desirable human factors of APL-style array programming can be implemented 
with competitive performance relative to more traditional techniques. This 
suggests that programs written in this highly abstract style may not suffer the 
performance gap traditionally assumed to go with their desirable features. 
This work is bolstered by previous work on the high-performance implemenation 
of array-oriented languages. 
\cite{ching1994experimental,ching1993primitive,
ching1990automatic,ju1991exploitation,
ju1991performance,bernecky1999reducing,schwarz1991acorn}

@@

The APEX compiler \cite{bernecky1997apex} developed vectorized approaches to handling
certain analyses to compile traditional APL, including a SIMD tokenizer
\cite{bernecky2003tokenizer}.
It also uses 
a matrix format to represent the AST. Traditional APL did not have
nested function definitions, however, and thus the APEX compiler does
not have any specific approaches to dealing with function lifting.

Timothy Budd implemented a compiler \cite{budd1984apl,budd2012apl} for APL which targeted
vector processors as well as C. Budd provided 
thoughts and some ideas on how the compiler might be implemented in
parallel as well.
